Finiteness in Cubical Type Theory

### 2020-07-01 sections/finiteness-predicates.tex
finished kuratowski finiteness section

In the case of \(x\;\AgdaFunction{\(\in\)}\;\mathit{xs}\) we can construct a
decision procedure for membership of a list, since we already have decidable
equality on the elements of the list, proving our obligation.
\section{Kuratowski Finiteness} \label{kuratowski} \todo{better intro}
The one big missing definition of finiteness to cover is \emph{Kuratowski}
finiteness.
While it's quite important, it's also quite different from the definitions we've
seen so far.
It starts with an encoding of the free join semilattice.
\begin{definition}[Free Join Semilattice]
  \(\mathcal{K}(A)\) is the free join semilattice, or, alternatively, the type
  of Kuratowski-finite subsets of \(A\).
  \begin{agdalisting}
    \ExecuteMetaData[agda/Algebra/Construct/Free/Semilattice/Definition.tex]{kuratowski-def}
  \end{agdalisting}

  We define it as a HIT.
  The first two constructors are point constructors, giving ways to create
  values of type \(\mathcal{K}(A)\).
  They are also recognisable as the two constructors for finite lists, a type
  which represents the free monoid.

  The next two constructors add extra paths to the type: equations that usage of
  the type must obey.
  These extra paths turn the free monoid into the free \emph{commutative} (com)
  \emph{idempotent} (dup) monoid.

  The final constructor enforces that the type \(\mathcal{K}(A)\) must be a set.
\end{definition}
The Kuratowski finite subset is a free join semilattice (or, equivalently, a
free commutative idempotent monoid).
More prosaically, \(\mathcal{K}\) is the abstract data type for finite sets, as
defined in the Boom hierarchy \cite{boomFurtherThoughtsAbstracto1981,
  bunkenburgBoomHierarchy1994}.
However, rather than just being a specification, \(\mathcal{K}\) is fully usable
as a data type in its own right, thanks to HITs.

Other definitions of \(\mathcal{K}\) exist (such as the one in
\cite{fruminFiniteSetsHomotopy2018}) which make the fact that \(\mathcal{K}\) is
the free join semilattice more obvious.
We have included such a definition in our formalisation, and proven it
equivalent to the one above.

Next, we need a way to say that an entire type is Kuratowski finite.
For that, we will need to define membership of \(\mathcal{K}\).
\begin{definition}[Membership of \(\mathcal{K}\)]
  Membership is defined by pattern-matching on \(\mathcal{K}\).
  The two point constructors are handled like so: \todo{to agda}
  \begin{equation}
    \begin{alignedat}{2}
      x \in&& \; []                      &\coloneqq \bot ; \\
      x \in&& \; y \dblcolon \mathit{ys} &\coloneqq \lVert x \equiv y \uplus x \in \mathit{ys} \rVert ;
    \end{alignedat}
  \end{equation}
  The \(\text{com}\) and \(\text{dup}\) constructors are handled by proving that
  the truncated form of \(\uplus\) is itself commutative and idempotent.
  The type of propositions is itself a set, satisfying the \(\text{trunc}\)
  constructor.
\end{definition}
Finally, we have enough background to define Kuratowski finiteness.
\begin{definition}[Kuratowski Finiteness] \phantom{break}
  \begin{agdalisting}
    \ExecuteMetaData[agda/Cardinality/Finite/Kuratowski.tex]{kuratowski-finite-def}
  \end{agdalisting}
\end{definition}

We also have the following two lemmas, proven in both
\cite{fruminFiniteSetsHomotopy2018} and our formalisation.
\begin{lemma}
  \(\mathcal{K}^f\) is a mere proposition.
\end{lemma}
\begin{lemma}
  This circle \(S^1\) is Kuratowski finite.
\end{lemma}
\subsection{Relation to Cardinal Finiteness}
\begin{lemma} \label{cardinal-kuratowski}
  Cardinal finiteness is equivalent to Kuratowski finiteness over a discrete
  set.
  \begin{equation}
    \mathcal{C}(A) \simeq \mathcal{K}^f(A) \times \text{Discrete}(A)
  \end{equation}
\end{lemma}
This proof is constructed by providing a pair of functions: one from
\(\mathcal{C}(A)\) to \(\mathcal{K}^f(A) \times \text{Discrete}(A)\), and one the
other way.
This pair implies an equivalence, because both source and target are
propositions.
The actual functions themselves are proven in our formalisation, as well as in
\cite{fruminFiniteSetsHomotopy2018}.

>>>>

In the case of \(x\;\AgdaFunction{\(\in\)}\;\mathit{xs}\) we can construct a
decision procedure for membership of a list, since we already have decidable
equality on the elements of the list, proving our obligation.
\section{Kuratowski Finiteness} \label{kuratowski}
We now finally arrive at the most important definition of finiteness: Kuratowski
finiteness.
As a definition, it is quite different from the predicates we've seen (it
doesn't involve lists, for instance), but it plays a much larger role in the
literature on finiteness predicates than, say, manifest enumerability.

We start with the definition of Kuratowski-finite subsets.
\begin{agdalisting}
  \ExecuteMetaData[agda/Algebra/Construct/Free/Semilattice/Definition.tex]{kuratowski-def}
\end{agdalisting}
The first two constructors are point constructors, giving ways to create
values of type \(\agdacal{K}\;A\).
They are also recognisable as the two constructors for finite lists, a type
which represents the free monoid.
The next two constructors add extra paths to the type: equations that usage of
the type must obey.
These extra paths turn the free monoid into the free \emph{commutative}
(\AgdaInductiveConstructor{com}) \emph{idempotent}
(\AgdaInductiveConstructor{dup}) monoid.
The final constructor truncates the type \(\agdacal{K}\;A\) to a set.

The Kuratowski finite subset is a free join semilattice (or, equivalently, a
free commutative idempotent monoid).
More prosaically, \(\agdacal{K}\) is the abstract data type for finite sets, as
defined in the Boom hierarchy \cite{boomFurtherThoughtsAbstracto1981,
  bunkenburgBoomHierarchy1994}.
However, rather than just being a specification, \(\agdacal{K}\) is fully usable
as a data type in its own right, thanks to HITs.

Other definitions of \(\agdacal{K}\) exist (such as the one in
\cite{fruminFiniteSetsHomotopy2018}) which make the fact that \(\agdacal{K}\) is
the free join semilattice more obvious.
We have included such a definition in our formalisation, and proven it
equivalent to the one above.

Next, we need a way to say that an entire type is Kuratowski finite.
For that, we will need to define membership of \(\agdacal{K}\).
\begin{equation}
  \begin{alignedat}{2}
    x\;\AgdaFunction{\ensuremath{\in}}& \; \AgdaInductiveConstructor{[]}                      &&= \agdabot \\
    x\;\AgdaFunction{\ensuremath{\in}}& \; y \;\AgdaInductiveConstructor{\ensuremath{\dblcolon}}\; \mathit{ys} &&= \AgdaDatatype{\ensuremath{\lVert}}\;\agdaequiv{x}{y}\;\AgdaDatatype{\ensuremath{\uplus}}\;x\;\AgdaFunction{\ensuremath{\in}}\;\mathit{ys}\;\AgdaDatatype{\ensuremath{\rVert}}
  \end{alignedat}
\end{equation}
The \AgdaInductiveConstructor{com} and \AgdaInductiveConstructor{dup}
constructors are handled by proving that the truncated form of \AgdaDatatype{\ensuremath{\uplus}}
itself commutative and idempotent.
The type of propositions is itself a set, satisfying the \AgdaInductiveConstructor{trunc}
constructor.
This gives us enough to define Kuratowski finiteness.
\begin{definition}[Kuratowski Finiteness]
  A type is Kuratowski finite if there exists a Kuratowski-finite subset of that
  type which contains every element of the type.
  \begin{agdalisting}
    \ExecuteMetaData[agda/Cardinality/Finite/Kuratowski.tex]{kuratowski-finite-def}
  \end{agdalisting}
\end{definition}

In our formalisation, we have proven a number of other lemmas regarding
Kuratowski finiteness, including the fact that
\AgdaFunction{\ensuremath{\mathcal{K}^f}} is a mere proposition, and that the
circle is Kuratowski finite.
These are also provided in \citet{fruminFiniteSetsHomotopy2018}, however, and
aren't directly relevant to what we're doing here so we won't provide the
details of the proof.
The key proof we will rely on for the rest of the paper is the following:
\begin{lemma} \label{cardinal-kuratowski}
  Cardinal finiteness is equivalent to Kuratowski finiteness over a discrete
  set.
  \begin{agdalisting}
    \ExecuteMetaData[agda/Cardinality/Finite/Kuratowski.tex]{card-iso-kuratowski}
  \end{agdalisting}
\end{lemma}
This proof is constructed by providing a pair of functions: one from
\(\mathcal{C}(A)\) to \(\agdacal{K}^f(A) \times \text{Discrete}(A)\), and one the
other way.
This pair implies an equivalence, because both source and target are
propositions. \todo{is it clear that this lemma is also in frumin et al?}
